---
title: "BIDS Format Support and DuckDB Backend"
author: "Yusuke Matsui"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIDS Format Support and DuckDB Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

PhysioIO provides two complementary features for managing collections of
physiological recordings:

1. **BIDS support** -- read and write data in the Brain Imaging Data
   Structure (BIDS) format, the community standard for organizing
   neuroimaging and electrophysiology datasets.
2. **DuckDB backend** -- store experiment metadata, channel information,
   events, and (optionally) signal data in an embedded analytical
   database for fast querying across many sessions.

This vignette demonstrates both capabilities with self-contained,
runnable examples that write to temporary locations.

```{r load}
library(PhysioIO)
```

We use a small deterministic object throughout.

```{r make-pe}
pe <- PhysioExperiment(
  assays = list(raw = matrix(0, nrow = 200, ncol = 3)),
  colData = S4Vectors::DataFrame(label = c("Fp1", "Fp2", "Cz")),
  samplingRate = 256
)
pe
```

## BIDS format support

### What is BIDS?

The Brain Imaging Data Structure (BIDS) is a standard for organizing
neuroimaging data into a consistent directory layout with machine-readable
metadata files. PhysioIO supports the BIDS-EEG and BIDS-iEEG extensions.

**Reference:** Gorgolewski KJ, et al. (2016). "The brain imaging data
structure, a format for organizing and describing outputs of neuroimaging
experiments." *Scientific Data*, 3, 160044.

### Writing to and reading from a BIDS dataset

`writeBIDS()` creates the required directory structure, writes the EDF
data file, and generates the sidecar TSV and JSON files. `readBIDS()`
loads it back, merging information from the companion sidecars.

```{r bids-roundtrip}
bids_root <- tempfile("bids")
dir.create(bids_root)

writeBIDS(pe, bids_root, subject = "01", task = "rest")

# Read a single recording back
pe_bids <- readBIDS(bids_root, subject = "01", task = "rest")
dim(pe_bids)
```

### Exploring and validating a BIDS dataset

```{r bids-explore}
# List all subjects
listBIDSSubjects(bids_root)

# Validate BIDS compliance
result <- validateBIDS(bids_root)
result$valid
result$n_subjects
```

The validator checks for `dataset_description.json`, subject directory
naming, modality directories, and required metadata fields.

```{r bids-cleanup}
unlink(bids_root, recursive = TRUE)
```

## DuckDB database backend

### Why use a database?

When a study involves dozens or hundreds of recording sessions, it
becomes impractical to load every file just to find which subjects had a
particular task or which recordings fall within a date range. The DuckDB
backend lets you register experiment metadata once and query it
efficiently.

**Reference:** Raasveldt M, Muehleisen H (2019). "DuckDB: an embeddable
analytical database." Proceedings of the 2019 International Conference on
Management of Data (SIGMOD).

The chunks below evaluate only when the optional `duckdb` package is
available.

### Connecting, initializing, registering, and querying

`initPhysioSchema()` creates the `experiments`, `channels`, `events`,
`signal_chunks`, `epochs`, and `annotations` tables.

```{r db-workflow, eval = requireNamespace("duckdb", quietly = TRUE)}
db_path <- tempfile(fileext = ".duckdb")
con <- connectDatabase(db_path)
initPhysioSchema(con)

# Register metadata (and optionally signal data)
registerExperiment(con, pe,
  experiment_id = "demo1",
  subject_id = "sub-01",
  task = "rest",
  store_signals = TRUE
)

# Query across sessions
all_exps <- queryExperiments(con)
sub01 <- queryExperiments(con, subject_id = "sub-01")

# Reload a registered experiment
pe_db <- loadExperiment(con, "demo1", load_signals = TRUE)
dim(pe_db)

# Summary statistics
stats <- dbStats(con)
stats$n_experiments

# Remove and disconnect
deleteExperiment(con, "demo1")
disconnectDatabase(con)
unlink(db_path)
```

## Session info

```{r sessioninfo}
sessionInfo()
```
