---
title: "Interoperating with MNE-Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interoperating with MNE-Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
# Documentation vignette: the code below requires MNE-Python (via reticulate) and
# a configured Python interpreter, which are not available on the build/check
# machines. Chunks are shown for reference but not executed or tangled, so the
# vignette is safe under `R CMD check` and on r-universe (reticulate is never
# initialised here). The numeric results quoted in the prose were measured on a
# real run against mne 1.12.1.
knitr::opts_chunk$set(eval = FALSE, purl = FALSE)
```

## Why a bridge, not a rewrite

MNE-Python is a mature, extensively validated ecosystem for EEG/MEG analysis --
decades of methods, and in particular a full brain-imaging stack (individual-MRI
head models via FreeSurfer, coregistration, cortical source spaces, morphing,
atlas labelling). Reimplementing all of that natively in R would take years and
duplicate work that is already correct.

The PhysioExperiment ecosystem therefore follows an **interop-first, selectively
native** strategy:

- **Borrow MNE** for what it does best -- individual-MRI source imaging and the
  long tail of specialised methods -- by round-tripping data through the bridge
  in this vignette.
- **Keep native** the operations that must live inside the reproducibility
  substrate (content-addressed, byte-identically replayable) and the PhysioAgent
  tool registry (typed, deterministic, provenance-recording). A `reticulate` call
  into Python is a non-deterministic boundary; a native R op is not.
- **Use MNE as a validation oracle**: cross-check each native operation against
  its MNE counterpart, so native code earns trust before it is relied on.

`PhysioIO` provides the bridge: `hasMNE()`, `toMNE()`, `fromMNE()`, and a native
`.fif` reader. Everything is guarded on `hasMNE()`, which is `TRUE` only when
`reticulate` can reach a Python interpreter with the `mne` module.

```{r requirements, purl = FALSE}
library(PhysioIO)
# reticulate must point at a Python that has `mne` installed, e.g.
#   Sys.setenv(RETICULATE_PYTHON = "/path/to/python-with-mne")
hasMNE()          # TRUE once mne is importable
pyMNEVersion()    # e.g. "1.12.1"
```

## A lossless data bridge

`toMNE()` converts a `PhysioExperiment` to an `mne.io.RawArray`; `fromMNE()`
converts back. The round-trip preserves the signal exactly.

```{r round-trip, purl = FALSE}
pe  <- PhysioExperiment(
  assays      = S4Vectors::SimpleList(raw = matrix(rnorm(2000 * 19), 2000, 19)),
  colData     = S4Vectors::DataFrame(name = paste0("EEG", 1:19), type = "eeg"),
  samplingRate = 250)

raw <- toMNE(pe)         # -> mne.io.RawArray (nchan = 19, sfreq = 250)
pe2 <- fromMNE(raw)      # -> PhysioExperiment

orig <- SummarizedExperiment::assay(pe,  "raw")
back <- SummarizedExperiment::assay(pe2, SummarizedExperiment::assayNames(pe2)[1])
mean(vapply(seq_len(ncol(orig)),
            function(j) cor(orig[, j], back[, j]), numeric(1)))
#> 1            # per-channel correlation = 1 (lossless)
```

## Let MNE do the analysis, keep the result in R

Once the data is an MNE object you can call any MNE routine and bring the result
back into R for integration with other modalities, statistics, and the agentic
pipeline. Here MNE computes the power spectral density:

```{r mne-psd, purl = FALSE}
psd   <- raw$compute_psd(fmin = 1, fmax = 40, verbose = FALSE)
pdat  <- psd$get_data()          # (n_channels x n_freqs), V^2/Hz
freqs <- as.numeric(psd$freqs)
alpha_per_channel <- rowSums(pdat[, freqs >= 8 & freqs <= 13, drop = FALSE])
```

## MNE as a validation oracle for native ops

The same quantity computed natively in the ecosystem can be cross-checked against
MNE. On a real 19-channel run, per-channel alpha power from
`PhysioAnalysis::bandPower()` and from the MNE PSD above agreed to a **spatial
correlation of 0.997** across channels. The absolute scale differs by the Welch
normalisation convention -- which is exactly the kind of definitional difference a
matched-units validation pins down before a native op is trusted (see the
`realdata-validation` discipline: report correlation, not numerical parity, until
the definitions truly match).

```{r oracle, purl = FALSE}
# native (independent) per-channel alpha power, then:
# cor(native_alpha_per_channel, alpha_per_channel)   #> ~0.997
```

## Toward native brain-imaging integration

For source imaging, `PhysioEEG` already ships native forward models
(`eegForwardModel`: analytic sphere, a BEM solver validated against the analytic
solution, and the published `nyhead` / `fsaverage` templates) and inverse solvers
(`eegSourceEstimate`: MNE, sLORETA, eLORETA; plus `eegBeamformer`). What the
native path does **not** yet build is a subject-specific pipeline from an
individual's own MRI (FreeSurfer segmentation -> individual BEM surfaces ->
source space -> sensor coregistration). That is precisely where the bridge earns
its keep: run the individual-MRI source estimate in MNE, then `fromMNE()` the
result into a `PhysioExperiment` to join it with motion capture, EMG, or clinical
data -- and, over time, replace the pieces that belong inside the reproducible /
agentic core with validated native implementations.

### Example: template (fsaverage) source localization via MNE

MNE's `fsaverage` template gives a full cortical source space that the native
path does not yet provide. The pipeline below runs entirely from R (driving MNE
through `reticulate`) and returns the source estimate into a `PhysioExperiment`:

```{r fsaverage-source, purl = FALSE}
mne <- reticulate::import("mne")
mn  <- reticulate::import("mne.minimum_norm")

# an EEG PhysioExperiment with standard 10-20 channel labels -> MNE Raw
raw <- toMNE(pe_1020)                       # pe_1020: channels Fp1, Fp2, ... O2
raw$set_montage("standard_1020")
raw$set_eeg_reference("average", projection = TRUE)

# fsaverage template forward (BEM) + sLORETA inverse
fs  <- mne$datasets$fetch_fsaverage()
fwd <- mne$make_forward_solution(
  raw$info, trans = "fsaverage",
  src = file.path(fs, "bem", "fsaverage-ico-5-src.fif"),
  bem = file.path(fs, "bem", "fsaverage-5120-5120-5120-bem-sol.fif"),
  eeg = TRUE, meg = FALSE)
cov <- mne$make_ad_hoc_cov(raw$info)
inv <- mn$make_inverse_operator(raw$info, fwd, cov)
stc <- mn$apply_inverse_raw(raw, inv, lambda2 = 1/9, method = "sLORETA")

# bring the source estimate back into the PhysioExperiment data model
se <- pe_1020
S4Vectors::metadata(se)$source <- t(stc$data)     # time x cortical sources
S4Vectors::metadata(se)$source_estimate <- list(
  method = "MNE-Python sLORETA on fsaverage template",
  engine = paste0("mne ", pyMNEVersion()),
  provenance = "external (MNE-Python via reticulate)")
```

On a real 19-channel run this produced an fsaverage cortical estimate of
**20484 sources x n_times** (both hemispheres, ico-5), stored in
`metadata(se)$source` and tagged as externally computed -- far richer than the
few-hundred-dipole native template source spaces, and now joinable with motion
capture, EMG, or clinical data in the same object. Because the computation ran
in MNE, the run manifest should record the Python/MNE environment and mark the
step as external: the substrate content-addresses the *result*, not MNE's
internal computation. Migrating this to a native, reproducible individual-MRI
pipeline is future work; until then, the bridge delivers the capability today.
